---
layout: post
title: "Geographic clustering of UK cities"
date: 2015-11-23 11:42:00 +0000
categories: clustering dbscan uk cities 
---

I know I am probably late to this party but I recently found out about DBSCAN or "A 
Density-Based Algorithm for Discovering Clusters in Large Spatial Databases with Noise"[^1]. In a nutshell, the algorithm visits successive data point and asks whether neighbouring points are density-reachable. In other words is it possible to connect two points with a chain of points all conforming to some density criteria. This has some major advantages over other clustering algorithms that I have used before.

* It can identify clusters of arbitrary shape.
* Number of clusters is not an input parameter.
* It's fast as it only visits the data points rather than the space in between.
* A data point with no close neighbours is assigned noise rather than its nearest cluster.

Let have a go at clustering uk cities from `library(maps)`. First load the packages and the data, then subset the data to get only the UK cities.

```{r, message = FALSE, warning = F}
library(ggplot2)
library(dplyr)
library(maps)
library(dbscan)

data("world.cities")
UK <- world.cities %>% filter(country.etc == "UK")
```

Now we can run the algorithm on the latitude and longitude collumns. Then we can pull the cluster assignments out of the resulting object.

```{r}
EPS <- 0.15
clusters <- dbscan(select(UK, lat, long), eps = EPS)
UK$cluster <- clusters$cluster
```

```{r, echo=FALSE}
theme_stripped <- theme(panel.background = element_blank(),
                        panel.grid = element_blank())
theme_empty <- theme(axis.text = element_blank(),
                     axis.title = element_blank(),
                     axis.ticks = element_blank())
```

Finally we can split the original data into two according to whether dbscan has assigned or cluster or noise.

```{r}
groups  <- UK %>% filter(cluster != 0)
noise  <- UK %>% filter(cluster == 0)
```

Now lets have a look at the results.

```{r}
ggplot(UK, aes(x = long, y = lat, alpha = 0.5)) + 
  geom_point(aes(fill = "grey"), noise) +
  geom_point(aes(colour = as.factor(cluster)), groups,
             size = 3) +
  coord_map() +
  theme_stripped +
  theme_empty +
  theme(legend.position = "none")
```

[1]: I recommend reading the paper which is quite accesible. Martin Ester, Hans-Peter Kriegel, Joerg Sander, Xiaowei Xu (1996). A Density-Based Algorithm for Discovering Clusters in Large Spatial Databases with Noise. Institute for Computer Science, University of Munich. Proceedings of 2nd International Conference on Knowledge Discovery and Data Mining (KDD-96).