---
title: "Downloading your twitter feed in R"
author: "Roman Popat"
date: "7 December 2015"
output: html_document
layout: post
---

The `twitteR` library is one of the most comprehensive R bindings for an API that I have ever seen. Thanks a million to [Jeff Gentry](https://github.com/geoffjentry/twitteR) for authoring and maintaining it. It took me a while to get to know it and I thought I'd share a little trick here. I recently wanted to trace the generation of my personal data from a morning in my life (possibly more on that later...). One aspect of this is the tweets that flow into my feed. `twitteR` has a very useful function for this.

```{r, message = FALSE, eval = F}
library(twitteR)
source("~/OneDriveBusiness/Projects/Authentications/twitterOAuth.R")
setup_twitter_oauth(consumer_key, consumer_secret, access_token, access_secret)
```

Now we're authenticated and ready to go, we can call `homeTimeline()` with the argument `retryOnRateLimit` set high. Note that twitter will only allow you to retrieve so many tweets from your feed so if you want recent tweets get them while you still can.

```{r, message = FALSE, eval = F}
timeLine <- homeTimeline(n = 800, retryOnRateLimit = 1000)
```

Now we can use some of the date functionality to subset the tweets retaining only those in the relevant time period.

```{r, message = FALSE, eval = F}
beginning <-  as.POSIXct("2015-12-07 09:00:00 UTC")
end <- as.POSIXct("2015-12-07 11:00:00 UTC")

times <- lapply(timeLine, function(x) x$created)
these <- which(times > beginning & times < end)

myMorningTweets <- timeLine[these]
```

Now that we have the subset that we want, we can use the provided function to turn the list like object into a data frame and store it for later.

```{r, message = FALSE, eval = F}
myMorningTweetsDF <- twListToDF(myMorningTweets)
write.csv(myMorningTweetsDF, "data/myMorningTweets.csv")
```


